Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-14 paper.tex
Foldable does not trivially implies rotations

The first step of diffusion process consists in applying a right
rotation over each individual column. However, we do not have enough
structure to identify a ``right'' or ``left'' direction over our
containers. The standard tool of the trade to achieve this effect is
the notion of foldable functor~\citep{yorgey:typeclassopedia}:
%
\coqfrom{Algebra}{Foldable}

Taking the monoid to be \coqe!list A! (under nil and concatenation,
\ie{} the free monoid construction), we can read off the
directionality we were after.
%
More usefully for our purposes, we can derive generic left and right
rotation operators for any foldable functor \coqe!T! Naperian
over a type \coqe!Ix!:
%
\coqfrom{Rows}{ror_sig}
\coqfrom{Rows}{rol_sig}

\TODO{Foldable to ror/rol}

>>>>

The first step of the diffusion process consists in applying a right
rotation over each individual column. However, we do not have enough
structure to identify a ``right'' or ``left'' direction over our
containers. To do so, we introduce the (regretfully ad-hoc\footnote{We
were hoping to be able to piggy-back on the notion of
foldable~\citep{yorgey:typeclassopedia} and Naperian functor to
identify directions. Intuitively, \coqe!fold! denotes a unique
left-to-right traversal order. However, we could only turn this into a
rotation if we asked for a decidable equality over the logarithm of
the functor. We are not sure yet whether we want to make such a
commitment.}) notion of circulant functor
%
\coqfrom{Algebra}{Circulant}
%
taking an \coqe!F!-vector to an \coqe!F! $\times$ \coqe!F! circulant
matrix (performing a right-rotation of \coqe!F A! at each step) and
\coqe!F! $\times$ \coqe!F! anticirculant matrix (performing a
left-rotation of \coqe!F A! at each step).

More usefully for our purposes, we can derive generic left and right
rotation operators for any circulant functor \coqe!T!  Naperian over a
type \coqe!Ix!:
%
\coqfrom{Algebra}{ror_rol_sig}

%% The equational theory is unclearas expected for rotations: additive on the
%% indices when going in the same direction (\coqe!ror i (ror j xs)! and
%% \coqe!rol i (rol j xs)!), subtractive when going in opposite direction
%% (\coqe!ror i (rol j xs)! and \coqe!rol i (ror j xs)!).

%\coqfrom{Algebra}{Foldable}
%% Taking the monoid to be \coqe!list A! (under nil and concatenation,
%% \ie{} the free monoid construction), we can read off the
%% directionality we were after.
%% %
